White Dwarf Mergers on Adaptive Meshes\\ II. Nuclear Reactions

### 2018-01-22 papers/wdmerger_II/paper.tex
Clean up formatting on density units

We terminate the simulation at $t = 9$ seconds, as we have found that for this
particular setup, all of the significant burning has completed by this time\footnote{In
some simulations we were mainly interested in seeing when the detonation happened, and
we terminated the simulation at that point. We show nucleosynthesis results only for
those simulations that had completed nuclear burning.}. At
this point, the total energy on the domain has become positive, typically about
$10^{51}$ ergs, and has stopped increasing. This means that the system has become
unbound due to nuclear energy release, and that no further meaningful nuclear energy
generation is occurring. In our discussions below, it helps to have some other
anchoring points in time when significant events occur. We arbitrarily define the
collision time, $t_{\rm c}$, as the time when the density at the center of the grid
first reaches $10^2\ \text{g / cm}^3$. This is reached only when the WDs meet at the
center of the domain. We define the detonation time, $t_{\rm d}$, as the time when the
temperature first reaches $4.5 \times 10^9\ \text{K}$. This is also an arbitrary
threshold, but we have found empirically that this temperature is only reached in
the collision when a detonation is formed.

As in Paper I, the region outside the WDs is filled with a low-density ambient
material with $\rho = 10^{-4}\ \text{g / cm}^{-3}$ whose composition and temperature is the
same as that of the WDs. Reactions in this region are unimportant, so for
computational efficiency, we disable all nuclear burning for zones that have
$\rho < 10^6\ \text{g / cm}^{-3}$ and $T < 10^8\ \text{K}$. We have confirmed that
the results do not depend on this choice. A sponge is applied to all gas of density
lower than $1\ \text{g / cm}^{-3}$. The sponge damps out the velocity in these zones
(see Paper I) to prevent spurious numerical error in low-density gas from dragging
down the global timestep. The density floor for the simulation is set
at $10^{-5}\ \text{g / cm}^{-3}$ and the temperature floor is $10^7$ K.

@@

Now let us examine the results for a collision of two 0.64 $\msolar$ WDs. We use
128 zones in the radial dimension, corresponding to a spatial resolution of 312.5 km.
Timesteps are limited only by the hydrodynamic stability constraint (and the plotfile
interval constraint). Resolutions in the range of a few hundred km have been a common
choice in earlier papers on white dwarf mergers and collisions. No adaptive mesh
refinement is used. For this configuration, the collision time is approximately $t_{\rm c} = 4.6\ \text{s}$,
and the detonation time is $t_{\rm d} = 7.43\ \text{s}$. The peak density at the detonation time
is $2.3 \times 10^7\ \text{g / cm}^3$. The detonation occurs at the center of the domain
where the two WDs initially meet.

@@

The upshot of this discussion is that the only way to resolve the
question is with higher resolution simulations. The semi-analytic
criterion provided by \cite{garg:2017}, and the numerical calculations
of \cite{seitenzahl:2009}, give critical radii for spontaneous detonation.
For the densities we are interested in ($\sim 10^7\ \text{g / cm}^3$),
assuming equal C/O material, the critical radii are in the range 1-10 km.
So, as pointed out by \citeauthor{kushnir:2013}, this numerical
stability issue is mainly a concern for simulations with
resolution of O(10 km) or worse. Detonations observed at higher
resolution are physically plausible because the critical temperature
gradient will be resolved. The high resolution needs to be achieved
with some combination of high resolution of the collision process
itself, and perhaps AMR based on the stability criterion as described
above, where we tag for refinement all zones above some $f_{s}$, the
ratio of the sound crossing time to the energy injection time. It
cannot be solely based on the burning stability criterion for the reason
described above, that a detonation may be spuriously locked in by
the low resolution regardless of how much resolution we add to it.

@@

At high resolution the collision time, $t_{\rm c} = 4.6\ \text{s}$, is similar to the
collision time at low resolution. However, the detonation time is around $t_{\rm d} = 6.62\ \text{s}$,
which is significantly earlier than at low resolution due to the high resolution simulations
being better able to resolve the point of contact between the WDs. The peak density crosses
$10^7\ \text{g / cm}^3$ significantly earlier, and this is what triggers the burning sufficient
to generate an earlier detonation. Approximately speaking, the later $t_{\rm d}$ is, the more
nickel is produced. As time passes, more white dwarf material slams into the center, compressing
and heating up the gas. Both effects are favorable to production of nickel. So, a later $t_{\rm d}$
implies that more material can be converted to nickel.

>>>>

We terminate the simulation at $t = 9$ seconds, as we have found that for this
particular setup, all of the significant burning has completed by this time\footnote{In
some simulations we were mainly interested in seeing when the detonation happened, and
we terminated the simulation at that point. We show nucleosynthesis results only for
those simulations that had completed nuclear burning.}. At
this point, the total energy on the domain has become positive, typically about
$10^{51}$ ergs, and has stopped increasing. This means that the system has become
unbound due to nuclear energy release, and that no further meaningful nuclear energy
generation is occurring. In our discussions below, it helps to have some other
anchoring points in time when significant events occur. We arbitrarily define the
collision time, $t_{\rm c}$, as the time when the density at the center of the grid
first reaches $10^2\ \text{g\,/\,cm}^3$. This is reached only when the WDs meet at the
center of the domain. We define the detonation time, $t_{\rm d}$, as the time when the
temperature first reaches $4.5 \times 10^9\ \text{K}$. This is also an arbitrary
threshold, but we have found empirically that this temperature is only reached in
the collision when a detonation is formed.

As in Paper I, the region outside the WDs is filled with a low-density ambient
material with $\rho = 10^{-4}\ \text{g\,/\,cm}^{3}$ whose composition and temperature is the
same as that of the WDs. Reactions in this region are unimportant, so for
computational efficiency, we disable all nuclear burning for zones that have
$\rho < 10^6\ \text{g\,/\,cm}^{3}$ and $T < 10^8\ \text{K}$. We have confirmed that
the results do not depend on this choice. A sponge is applied to all gas of density
lower than $1\ \text{g\,/\,cm}^{3}$. The sponge damps out the velocity in these zones
(see Paper I) to prevent spurious numerical error in low-density gas from dragging
down the global timestep. The density floor for the simulation is set
at $10^{-5}\ \text{g\,/\,cm}^{3}$ and the temperature floor is $10^7$ K.

@@

Now let us examine the results for a collision of two 0.64 $\msolar$ WDs. We use
128 zones in the radial dimension, corresponding to a spatial resolution of 312.5 km.
Timesteps are limited only by the hydrodynamic stability constraint (and the plotfile
interval constraint). Resolutions in the range of a few hundred km have been a common
choice in earlier papers on white dwarf mergers and collisions. No adaptive mesh
refinement is used. For this configuration, the collision time is approximately $t_{\rm c} = 4.6\ \text{s}$,
and the detonation time is $t_{\rm d} = 7.43\ \text{s}$. The peak density at the detonation time
is $2.3 \times 10^7\ \text{g\,/\,cm}^3$. The detonation occurs at the center of the domain
where the two WDs initially meet.

@@

The upshot of this discussion is that the only way to resolve the
question is with higher resolution simulations. The semi-analytic
criterion provided by \cite{garg:2017}, and the numerical calculations
of \cite{seitenzahl:2009}, give critical radii for spontaneous detonation.
For the densities we are interested in ($\sim 10^7\ \text{g\,/\,cm}^3$),
assuming equal C/O material, the critical radii are in the range 1-10 km.
So, as pointed out by \citeauthor{kushnir:2013}, this numerical
stability issue is mainly a concern for simulations with
resolution of O(10 km) or worse. Detonations observed at higher
resolution are physically plausible because the critical temperature
gradient will be resolved. The high resolution needs to be achieved
with some combination of high resolution of the collision process
itself, and perhaps AMR based on the stability criterion as described
above, where we tag for refinement all zones above some $f_{s}$, the
ratio of the sound crossing time to the energy injection time. It
cannot be solely based on the burning stability criterion for the reason
described above, that a detonation may be spuriously locked in by
the low resolution regardless of how much resolution we add to it.

@@

At high resolution the collision time, $t_{\rm c} = 4.6\ \text{s}$, is similar to the
collision time at low resolution. However, the detonation time is around $t_{\rm d} = 6.62\ \text{s}$,
which is significantly earlier than at low resolution due to the high resolution simulations
being better able to resolve the point of contact between the WDs. The peak density crosses
$10^7\ \text{g\,/\,cm}^3$ significantly earlier, and this is what triggers the burning sufficient
to generate an earlier detonation. Approximately speaking, the later $t_{\rm d}$ is, the more
nickel is produced. As time passes, more white dwarf material slams into the center, compressing
and heating up the gas. Both effects are favorable to production of nickel. So, a later $t_{\rm d}$
implies that more material can be converted to nickel.
